---
title: "Day 1 lab solutions"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("dplyr")
```

Read in the msleep_ggplot2.csv file with the function read.csv and use the function class to determine what type of object is returned.


```{r}
data(sleep)
class(sleep)
tlp <- read.csv('https://raw.githubusercontent.com/tidyverse/ggplot2/master/data-raw/msleep.csv')
class(tlp)
```

Now use the filter function to select only the primates. How many animals in the table are primates? Hint: the nrow function gives you the number of rows of a data frame or matrix.

```{r}
onlyPrimates <- filter(tlp, order=='Primates')
nrow(onlyPrimates)
```

What is the class of the object you obtain after subsetting the table to only include primates?

```{r}
class(onlyPrimates)
```

Now use the select function to extract the sleep (total) for the primates. What class is this object? Hint:  to pipe the results of the filter function to select.

```{r}
filter(tlp,order=='Primates') %>%
    select(sleep_total) -> sleepPrimates
select(filter(tlp,order=='Primates'), sleep_total) #equivalent
class(sleepPrimates)
```

Now we want to calculate the average amount of sleep for primates (the average of the numbers computed above). One challenge is that the mean function requires a vector so, if we simply apply it to the output above, we get an error. Look at the help file for unlist and use it to compute the desired average.

```{r}
filter(tlp,order=='Primates') %>%
    select(sleep_total) %>%
    unlist %>%
    mean
mean(unlist(sleepPrimates)) #equivalent
```

For the last exercise, we could also use the dplyr summarize function. We have not i#ntroduced this function, but you can read the help file and repeat exercise 5, this time using just filter and summarize to get the answer.

```{r}
tlp %>%
  filter(order=='Primates') %>%
    summarise(., mean(sleep_total))
```