---
title: "Population, Samples, and Estimates Exercises"
author: "Alexis Marchesini"
date: "23/02/2016"
output:  html_document
---

# Population, Samples, and Estimates Exercises

### Downloading external input file and importing it in R

```{r}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- basename(url)
download(url, destfile=filename)
mice_pheno <- read.csv(filename)
```
We will remove the lines that contain missing values:
```{r}
mice_pheno <- na.omit(mice_pheno)
```

### Question 1

Use `dplyr` to create a vector `x` with the body weight of all males on the control (`chow`) `diet`. What is this population’s average?

```{r}
library(dplyr)
x <- filter(mice_pheno,Sex == "M" & Diet == "chow") %>% select(Bodyweight) %>% unlist
mean_x=mean(x)
mean_x
```

### Question 2

Now use the `rafalib` package and use the `popsd` function to compute the population standard deviation.
```{r}
library(rafalib)
sd_x=popsd(x)
sd_x
```
### Question 3
Set the seed at 1. Take a random sample X of size 25 from `x`. What is the sample average?
```{r}
set.seed(1) 
X=sample(x,25)
mean_X=mean(X)
mean_X
```
### Question 4
Use dplyr to create a vector `y` with the body weight of all males on the high fat (`hf`) diet. What is this population’s average?
```{r}
y <- filter(mice_pheno,Sex == "M" & Diet == "hf") %>% select(Bodyweight) %>% unlist
mean_y=mean(y)
mean_y
```
### Question 5
Now use the `rafalib` package and use the `popsd` function to compute the population standard deviation.

```{r}
library(rafalib)
sd_y=popsd(y)
sd_y
```
### Question 6
Set the seed at 1. Take a random sample Y of size 25 from y. What is the sample average?

```{r}
set.seed(1) 
Y=sample(y,25)
mean_Y=mean(Y)
mean_Y
```

### Question 7
What is the difference in absolute value between y¯−x¯ and X¯ - Y¯ ?
```{r}
sample_mean_diff=abs(mean_Y-mean_X)
sample_mean_diff
pop_mean_diff= abs(mean_y-mean_x)
pop_mean_diff
sample_pop_mean_diff= abs(pop_mean_diff-sample_mean_diff)
sample_pop_mean_diff
```
### Question 8
Repeat the above for females. Make sure to set the seed to 1 before each sample call.
What is the difference in absolute value between y¯−x¯ and X¯ - Y¯ ?

```{r}
library(dplyr)
xF <- filter(mice_pheno,Sex == "F" & Diet == "chow") %>% select(Bodyweight) %>% unlist
mean_xF=mean(xF)
library(rafalib)
sd_xF=popsd(xF)
sd_xF

set.seed(1) 
XF=sample(xF,25)
mean_XF=mean(XF)
mean_XF

yF <- filter(mice_pheno,Sex == "F" & Diet == "hf") %>% select(Bodyweight) %>% unlist
mean_yF=mean(yF)
library(rafalib)
sd_yF=popsd(yF)
sd_yF

set.seed(1) 
YF=sample(yF,25)
mean_YF=mean(YF)
mean_YF

sample_mean_diff_F=abs(mean_YF-mean_XF)
sample_mean_diff_F
pop_mean_diff_F= abs(mean_yF-mean_xF)
pop_mean_diff_F

sample_pop_mean_diff_F= abs(pop_mean_diff_F-sample_mean_diff_F)
sample_pop_mean_diff_F
```
### Question 9
For the females, our sample estimates were closer to the population difference than with males. What is a possible explanation for this?
- A) The population variance of the females is smaller than that of the males; thus, the
sample variable has less variability.
- B) Statistical estimates are more precise for females.
- C) The sample size was larger for females.
- D) The sample size was smaller for females.

**Answer**: **A**

### License
PH525x, Rafael Irizarry and Michael Love, MIT License 

The **rmarkdown** package is licensed under the GPLv3 (http://www.gnu.org/licenses/gpl.html).