---
title: "MZ_d2_e2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Use the Rmd script to load the spider dataset. Suppose we build a model using two variables: ~ type + leg.

```{r, echo=FALSE}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
library(downloader)
if (!file.exists(filename))
    download(url, filename)
spiders <- read.csv(filename, skip=1)
```

```{r}
#spiders <- read.csv('spiders.csv') #download the spiders dataset or give instructions, to make the script reproducible
```

What is the t-statistic for the contrast of leg pair L4 vs. leg pair L2?

```{r}
spiders <- within(spiders, leg <- relevel(leg,"L2"))
spidersLM1 <- lm(friction ~ type + leg,data=spiders)
print(summary(spidersLM1))
```
Suppose that we notice that the within-group variances for the groups with smaller frictional coefficients are generally smaller, and so we try to apply a transformation to the frictional coefficients to make the within-group variances more constant.

Add a new variable log2friction to the spider dataframe:

```{r}
spiders$log2friction <- log2(spiders$friction)
boxplot(log2friction ~ type*leg, data=spiders)
```

Run a linear model of log2friction with type, leg and interactions between type and leg.

```{r}
spiders <- within(spiders, leg <- relevel(leg,"L1"))
spidersLM2 <- lm(log2friction ~ type*leg,data=spiders)
print(summary(spidersLM2))

```

What is the t-statistic for the interaction of type push and leg L4? 

```{r}
print(summary(spidersLM2)$coeff[,"t value"]['typepush:legL4'])
```

What is the L2 vs. L1 estimate in log2(friction) for the pull samples?

(Idea #1 filter for "pull" from the original dataset)

```{r}
library('dplyr')
pullingSpiders <- filter(spiders,type=='pull')
pullingSpiders <- within(pullingSpiders, leg <- relevel(leg,"L2"))

spidersLM3 <- lm(log2friction ~ leg,data=pullingSpiders)
print (summary(spidersLM3))

print (spidersLM3$coefficients['legL1'])
```


(Idea #2)
```{r}
library('dplyr') 
spiders <- within(spiders, leg <- relevel(leg,"L2"))

spidersLM3b <- lm(log2friction ~ leg*type,data=spiders)
print (summary(spidersLM3b))

print (spidersLM3b$coefficients['legL1'])
```

What is the L2 vs. L1 estimate in log2(friction) for the push samples? Remember, because of the interaction terms, this is not the same as the L2 vs L1 difference for the pull samples. If you’re not sure use the contrast function. Another hint: consider the arrows plot for the model with interactions.

>> Shall we evaluate "legL1:typepush" with "log2friction ~ leg * type" ?

```{r}

spiders <- within(spiders, leg <- relevel(leg,"L2"))

spidersLM4 <- lm(log2friction ~ leg*type,data=spiders)
print (summary(spidersLM4))

print (spidersLM4$coefficients['legL1'])
```
