---
title: "Population, Samples, and Estimates Exercises"
author: "Francesco Asnicar"
date: "22 February 2016"
output: html_document
---

For these exercises, we will be using the following dataset:
```{r}
library(downloader) 
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- basename(url)
download(url, destfile=filename)
dat <- read.csv(filename)
```

We will remove the lines that contain missing values:
```{r}
dat <- na.omit(dat)
```

# 1
What is this population's average?
```{r}
library(dplyr)
x <- dplyr::filter(dat, dat$Sex=='M', dat$Diet=='chow')
pop_avg <- mean(x$Bodyweight)
pop_avg
```

# 2
Compute the population standard deviation.
```{r}
library(rafalib)
pop_sd <- rafalib::popsd(x$Bodyweight)
pop_sd
```

# 3
What is the sample average?
```{r}
set.seed(1)
X <- sample(x$Bodyweight, size=25)
bar_X <- mean(X)
bar_X
```
# 4
What is this population's average?
```{r}
y <- dplyr::filter(dat, dat$Sex=='M', dat$Diet=='hf')
m_hf_avg <- mean(y$Bodyweight)
m_hf_avg
```

# 5
Compute the population standard devitian.
```{r}
m_hf_sd <- rafalib::popsd(y$Bodyweight)
m_hf_sd
```

# 6
What is the sample average?
```{r}
set.seed(1)
Y <- sample(y$Bodyweight, size=25)
bar_Y <- mean(Y)
bar_Y
```

# 7
What is the difference in absolute value for the population
```{r}
abs(pop_avg-m_hf_avg)
```

and sample
```{r}
abs(bar_X-bar_Y)
```

# 8
What is the difference in absolute value for the population
```{r}
female_chow <- dplyr::filter(dat, dat$Sex=='F', dat$Diet=='chow')
female_hf <- dplyr::filter(dat, dat$Sex=='F', dat$Diet=='hf')

pop_chow_avg <- mean(female_chow$Bodyweight)
pop_chow_sd <- rafalib::popsd(female_chow$Bodyweight)

pop_hf_avg <- mean(female_hf$Bodyweight)
pop_hf_sd <- rafalib::popsd(female_hf$Bodyweight)

set.seed(1)
sample_chow <- sample(female_chow$Bodyweight, size=25)

set.seed(1)
sample_hf <- sample(female_hf$Bodyweight, size=25)

sample_chow_avg <- mean(sample_chow)
sample_chow_sd <- rafalib::popsd(sample_chow)

sampe_hf_avg <- mean(sample_hf)
sampe_hf_sd <- rafalib::popsd(sample_hf)

abs(pop_hf_avg-pop_chow_avg)
```

and sample
```{r}
abs(sample_chow_avg-sampe_hf_avg)
```

# 9
What is a possible explanation for this?

A) The population variance of the females is smaller than that of the males; thus, the sample variable has less variability.

Control
```{r}
pop_chow_sd
sample_chow_sd
```

High fat
```{r}
pop_hf_sd
sampe_hf_sd
```