---
title: "Applied Statistics for High-throughput Biology: Session 4"
output: slidy_presentation
author: "Levi Waldron"
vignette: >
  %\VignetteIndexEntry{Applied Statistics for High-throughput Biology: Session 4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
library(AppStatBio)
```

## Outline
- [Directed acyclic graphs](https://pmc.ncbi.nlm.nih.gov/articles/PMC8821727/)
- [Exploratory data analysis](https://genomicsclass.github.io/book/pages/exploratory_data_analysis.html) (Chapter 2)
    - Quantile Quantile (QQ)-plots and regression diagnostics
    - Volcano plots
    - MA plots
    - heatmaps
- [Batch effects](http://genomicsclass.github.io/book/pages/intro_to_batch_effects.html) (Chapter 10)


```{r dag_layout, include=FALSE}
dag_layout <- list(
  x = c(I.instrumental_variable = -2, E.exposure = 0, C.confounder = -1, G.measured_confounder = 2, M.mediator = 2, S.collider = 2, L.unmeasured_cause = 2, D.disease = 4, J.effect_modifier = 4, Z.descendant = 6),
  y = c(I.instrumental_variable = 0, E.exposure = 0, C.confounder = 2, G.measured_confounder = 2, M.mediator = 1, S.collider = -1, L.unmeasured_cause = -2.5, D.disease = 0, J.effect_modifier = 2, Z.descendant = 0)
)
```

## Causal Directed Acyclic Graphs

Digitale, J. C., Martin, J. N., & Glymour, M. M. (2022). _Tutorial on directed acyclic graphs._ Journal of clinical epidemiology, 142, 264–267. [https://doi.org/10.1016/j.jclinepi.2021.08.001](https://doi.org/10.1016/j.jclinepi.2021.08.001)

- **Paths** are sequences of arrows, of any direction, connecting two variables and may be causal or non-causal.
- **Paths are causal** if each variable causes the subsequent variable (all the arrows point in the same direction).
- **Paths are non-causal** if the arrows do not all point in the same direction. They contain confounders and/or colliders.

```{r dag_basic, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
library(ggdag)
library(ggplot2)

dag_basic <- dagify(
  D.disease ~ E.exposure,
  coords = dag_layout
)
ggplot(dag_basic, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node(color = "lightgrey") +
  geom_dag_text_repel(aes(label = name), color = "black", show.legend = FALSE) +
  theme_dag()
```

## Adding Confounders

- **Confounding** occurs because of common (shared) causes (e.g., C) of E and D. To estimate the effect of E on D, it is necessary to control for such common causes or other variables along the non-causal path. Control for either C or G would be adequate to eliminate the confounding due to C. G may be preferable, for example, if it is easier to obtain a high-quality measurement of G.

```{r dag_confounder, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
dag_confounder <- dagify(
  D.disease ~ E.exposure + G.measured_confounder,
  E.exposure ~ C.confounder,
  G.measured_confounder ~ C.confounder,
  coords = dag_layout
)
ggplot(dag_confounder, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node(color = "lightgrey") +
  geom_dag_text_repel(aes(label = name), color = "black", show.legend = FALSE) +
  theme_dag()
```

## Adding Mediators

- **Mediators** (e.g., M) are caused by E and, in turn, cause D. They should not be controlled for to estimate the total effect of E on D. (Control for them **only** if you want to estimate the **direct** effect of E on D, excluding the mediated pathway).

```{r dag_mediator, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
dag_mediator <- dagify(
  D.disease ~ E.exposure + G.measured_confounder + M.mediator,
  E.exposure ~ C.confounder,
  G.measured_confounder ~ C.confounder,
  M.mediator ~ E.exposure,
  coords = dag_layout
)
ggplot(dag_mediator, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node(color = "lightgrey") +
  geom_dag_text_repel(aes(label = name), color = "black", show.legend = FALSE) +
  theme_dag()
```

## Adding Colliders

- **Colliders** (e.g., S) are so named because they have two arrows pointing into them. Colliders on a path block that path unless they are conditioned on (e.g., by controlling for them) or a consequence of the collider is conditioned on.

```{r dag_collider, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
dag_collider <- dagify(
  D.disease ~ E.exposure + G.measured_confounder + M.mediator + L.unmeasured_cause,
  E.exposure ~ C.confounder,
  G.measured_confounder ~ C.confounder,
  M.mediator ~ E.exposure,
  S.collider ~ E.exposure + L.unmeasured_cause,
  coords = dag_layout
)
ggplot(dag_collider, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node(color = "lightgrey") +
  geom_dag_text_repel(aes(label = name), color = "black", show.legend = FALSE) +
  theme_dag()
```

## Adding Descendants

- **Descendants** of D (e.g., Z). Analyses should not adjust for, stratify on, or in any way condition on descendants.

```{r dag_descendant, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
dag_descendant <- dagify(
  D.disease ~ E.exposure + G.measured_confounder + M.mediator + L.unmeasured_cause,
  E.exposure ~ C.confounder,
  G.measured_confounder ~ C.confounder,
  M.mediator ~ E.exposure,
  S.collider ~ E.exposure + L.unmeasured_cause,
  Z.descendant ~ D.disease,
  coords = dag_layout
)
ggplot(dag_descendant, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node(color = "lightgrey") +
  geom_dag_text_repel(aes(label = name), color = "black", show.legend = FALSE) +
  theme_dag()
```

## Adding Instrumental Variables

- **Instrumental variables** (e.g., I) are variables related to the exposure of interest that have no association with the outcome except through the exposure. Instrumental variables analysis can be used to derive effect estimates when there is intractable confounding of E and D.

```{r dag_instrument, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
dag_instrument <- dagify(
  D.disease ~ E.exposure + G.measured_confounder + M.mediator + L.unmeasured_cause,
  E.exposure ~ C.confounder + I.instrumental_variable,
  G.measured_confounder ~ C.confounder,
  M.mediator ~ E.exposure,
  S.collider ~ E.exposure + L.unmeasured_cause,
  Z.descendant ~ D.disease,
  coords = dag_layout
)
ggplot(dag_instrument, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node(color = "lightgrey") +
  geom_dag_text_repel(aes(label = name), color = "black", show.legend = FALSE) +
  theme_dag()
```

## The Full DAG

- **Effect modifiers** (e.g., J) are variables that cause D and modify the effect of other causes of D, such as E. If E and J both cause D, then J modifies the effect of E on D on at least one effect-measure scale (additive or multiplicative).

```{r dag1, echo=FALSE, fig.width=10, fig.height=6, warning=FALSE, message=FALSE}
library(ggdag)
library(ggplot2)

dag <- dagify(
  D.disease ~ E.exposure + G.measured_confounder + M.mediator + L.unmeasured_cause + J.effect_modifier,
  E.exposure ~ C.confounder + I.instrumental_variable,
  G.measured_confounder ~ C.confounder,
  M.mediator ~ E.exposure,
  S.collider ~ E.exposure + L.unmeasured_cause,
  Z.descendant ~ D.disease,
  coords = dag_layout
)

ggplot(dag, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_node(color = "lightgrey") +
  geom_dag_text_repel(aes(label = name), color = "black", show.legend = FALSE) +
  theme_dag()
```

## Key Concepts

### General
- **Paths** are sequences of arrows, of any direction, connecting two variables and may be causal or non-causal.
- **Paths are causal** if each variable causes the subsequent variable (all the arrows point in the same direction).
- **Paths are non-causal** if the arrows do not all point in the same direction. They contain confounders and/or colliders.
- **Confounding** occurs because of common (shared) causes (e.g., C) of E and D. To estimate the effect of E on D, it is necessary to control for such common causes or other variables along the non-causal path. Control for either C or G would be adequate to eliminate the confounding due to C. G may be preferable, for example, if it is easier to obtain a high-quality measurement of G.
- **Effect modifiers** (e.g., J) are variables that cause D and modify the effect of other causes of D, such as E. If E and J both cause D, then J modifies the effect of E on D on at least one effect-measure scale (additive or multiplicative).

### Do NOT control for
- **Mediators** (e.g., M) are caused by E and, in turn, cause D. They should not be controlled for to estimate the total effect of E on D. (Control for them **only** if you want to estimate the **direct** effect of E on D, excluding the mediated pathway).
- **Colliders** (e.g., S) are so named because they have two arrows pointing into them. Colliders on a path block that path unless they are conditioned on (e.g., by controlling for them) or a consequence of the collider is conditioned on.
- **Descendants** of D (e.g., Z). Analyses should not adjust for, stratify on, or in any way condition on descendants.
- **Instrumental variables** (e.g., I) are variables related to the exposure of interest that have no association with the outcome except through the exposure. Instrumental variables analysis can be used to derive effect estimates when there is intractable confounding of E and D.

# Exploratory data analysis

## Introduction

> “The greatest value of a picture is when it forces us to notice what we never expected to see.” - John W. Tukey

- Discover biases, systematic errors and unexpected variability in data.
- Graphical approach to detecting these issues. Represents a first step in data analysis and guides hypothesis testing.
- EDA helps us check the assumptions of our statistical tests.
- Opportunities for discovery are often in the outliers.

## Quantile Quantile Plots

- **Why use them?** A primary tool for checking if our data follows a theoretical distribution.`
- Quantiles divide a distribution into equally sized bins (e.g., 100 bins for percentiles).
- We plot the quantiles from our data against the theoretical quantiles of a distribution (e.g., the normal distribution).
- If our data perfectly matches the theoretical distribution, the points will form a straight line ($y=x$). Deviations from the line indicate our data does not fit that distribution.

## Example: Quantile Quantile Plots

```{r qq1, echo=FALSE, fig.width=10, fig.height=4, warning=FALSE}
suppressPackageStartupMessages({
  library(UsingR)
  library(rafalib)
  library(ggplot2)
  library(patchwork)
})

# height qq plot
p1 <- ggplot(data.frame(x = father.son$fheight), aes(sample = x)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Height Q-Q Plot", x = "Theoretical Normal Quantiles", y = "Height Quantiles") +
  theme_minimal()

# t-distributed for 12 df
p2 <- ggplot(data.frame(x = rt(1000, 12)), aes(sample = x)) +
  stat_qq(distribution = qt, dparams = list(df = 12)) +
  stat_qq_line(distribution = qt, dparams = list(df = 12)) +
  labs(title = "T Quantiles (df=12) Q-Q Plot", x = "Theoretical t quantiles", y = "Sample Quantiles") +
  coord_cartesian(ylim = c(-6, 6)) +
  theme_minimal()

# t-distributed for 3 df
p3 <- ggplot(data.frame(x = rt(1000, 3)), aes(sample = x)) +
  stat_qq(distribution = qt, dparams = list(df = 3)) +
  stat_qq_line(distribution = qt, dparams = list(df = 3)) +
  labs(title = "T Quantiles (df=3) Q-Q Plot", x = "Theoretical t quantiles", y = "Sample Quantiles") +
  coord_cartesian(ylim = c(-6, 6)) +
  theme_minimal()

p1 | p2 | p3
```

## Regression Diagnostics

```{r reg_diag, echo=FALSE, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
fit <- lm(dist ~ speed, data = cars)
par(mfrow = c(2, 2))
plot(fit)
```

## Exploratory data analysis in high dimensions


```{r eda1}
library(GSE5859Subset)
data(GSE5859Subset) ## this loads three tables
class(geneExpression)
dim(geneExpression)
class(sampleInfo)
dim(sampleInfo)
head(sampleInfo)
```

## Volcano plots: Setup

T-test for every row (gene) of gene expression matrix:
```{r vp1, message=FALSE}
library(genefilter)
g <- factor(sampleInfo$group)
system.time(results <- rowttests(geneExpression, g))
pvals <- results$p.value
```

Note that these 8,793 tests are done in about 0.01s

## Volcano plots: Example

```{r vp2, echo = FALSE, fig.height=5, fig.width=5, warning=FALSE}
library(ggplot2)
library(dplyr)

bonferroni_thresh <- -log10(0.05 / nrow(geneExpression))

df_volcano <- data.frame(
  EffectSize = results$dm,
  NegLogP = -log10(results$p.value)
) %>%
  mutate(Significant = NegLogP > bonferroni_thresh)

ggplot(df_volcano, aes(x = EffectSize, y = NegLogP, color = Significant)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_hline(yintercept = bonferroni_thresh, color = "red", linetype = "dashed") +
  scale_color_manual(values = c("black", "red")) +
  labs(x = "Effect size (difference in group means)", y = "-log10(p-value)") +
  theme_minimal() +
  theme(legend.position = "none") +
  annotate("text",
    x = max(df_volcano$EffectSize), y = bonferroni_thresh + 0.2,
    label = "Bonferroni = 0.05", color = "red", hjust = 1
  )
```

## Volcano plots: Summary

- A volcano plot lets us visualize both the **statistical significance** (p-value) and **biological significance** (effect size or fold change) at the same time for thousands of genes.
- **Top-right/left corners:** Genes with large effect sizes and high statistical significance. These are often the most interesting candidates.
- **Top-center:** Genes that are statistically significant but have a small effect size.
- **Bottom:** Genes that are not statistically significant, regardless of their effect size.
- Can color points by significance threshold. Check for asymmetry, which might indicate biases.

## P-value histograms: Setup

- If all null hypotheses are true (i.e., no genes are truly differentially expressed), we expect a **flat histogram** of p-values, where every p-value from 0 to 1 is equally likely.

```{r pvalhist1}
m <- nrow(geneExpression)
n <- ncol(geneExpression)
set.seed(1)
randomData <- matrix(rnorm(n * m), m, n)
nullpvals <- rowttests(randomData, g)$p.value
```

## P-value histograms: Example

```{r pvalhist2, echo=FALSE, fig.height=3, fig.width=8, warning=FALSE}
library(ggplot2)
library(patchwork)

df_pvals <- data.frame(
  Original = pvals,
  Random = nullpvals
)

p1 <- ggplot(df_pvals, aes(x = Original)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "gray", color = "black") +
  coord_cartesian(ylim = c(0, 1400)) +
  labs(title = "Original Data p-values", x = "p-value", y = "Frequency") +
  theme_minimal()

p2 <- ggplot(df_pvals, aes(x = Random)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "gray", color = "black") +
  coord_cartesian(ylim = c(0, 1400)) +
  labs(title = "Random Data p-values", x = "p-value", y = "Frequency") +
  theme_minimal()

p1 | p2
```

## P-value histograms: Example 2 (permutation)

Note that permuting these data doesn't produce an ideal null p-value histogram due to batch effects:

```{r pvalhist3, fig.height=3, fig.width=4, warning=FALSE}
permg <- sample(g)
permresults <- rowttests(geneExpression, permg)

ggplot(data.frame(pvalue = permresults$p.value), aes(x = pvalue)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "gray", color = "black") +
  labs(title = "Permuted Data p-values", x = "p-value", y = "Frequency") +
  theme_minimal()
```

## P-value histograms: Summary

- Give a quick look at the overall results of a high-throughput experiment. A spike near zero suggests the presence of differentially expressed genes.
- A non-uniform histogram for permuted data is a red flag, suggesting non-independence between samples, often due to hidden batch effects.

## MA plot

- **Why use it?** An MA plot is a clever transformation of a scatter plot, designed to better visualize differences between two samples (or one sample and a reference). It's just a scatterplot rotated 45$^o$.
- The rotation helps us see systematic biases. The 'A' (Average) on the x-axis represents overall signal intensity, and the 'M' (Minus, or log-ratio) on the y-axis represents the fold change. This makes it much easier to see if the fold change is dependent on gene intensity.

```{r pvalhist4, fig.height=4, fig.width=8, warning=FALSE}
library(ggplot2)
library(patchwork)
library(dplyr)

pseudo <- apply(geneExpression, 1, median)
df_ma <- data.frame(
  Sample = geneExpression[, 1],
  Pseudo = pseudo
) %>%
  mutate(
    A = (Sample + Pseudo) / 2,
    M = Sample - Pseudo
  )

p1 <- ggplot(df_ma, aes(x = Pseudo, y = Sample)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Standard scatter plot", x = "Median across samples", y = "Sample 1") +
  theme_minimal()

p2 <- ggplot(df_ma, aes(x = A, y = M)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "MA plot", x = "Average (A)", y = "Minus (M)") +
  theme_minimal()

p1 | p2
```

## MA plot: Summary

- Useful for quality control of high-dimensional data.
- In an ideal MA plot, the cloud of points is centered on y=0 with no trend.
- `affyPLM::MAplots` creates better MA plots, adding a smoothing line to highlight departures from the horizontal.

## Heatmaps

* Detailed representation of a high-dimensional dataset. The `ComplexHeatmap` package is the best as of 2025.
* **Important Note:** Before plotting, we usually **scale** the data for each gene. This ensures the color pattern is driven by relative expression changes, not by a few highly expressed genes dominating the color scale.

```{r ma1, fig.width=12, echo=FALSE}
suppressPackageStartupMessages(library(ComplexHeatmap))
keep <- rank(apply(geneExpression, 1, var)) <= 100
ge <- geneExpression[keep, ]
ge <- t(scale(t(ge))) # Scale genes across samples
rownames(ge) <- NULL
colnames(ge) <- NULL
chr <- sub("chr", "", geneAnnotation$CHR)
chr[is.na(chr)] <- "Un"
chr <- factor(chr, levels = c(1:22, "Un", "X", "Y"))
chrcols <- list(chromosome = c(colorRampPalette(c("lightgrey", "black"))(22), "white", "pink", "blue"))
names(chrcols[["chromosome"]]) <- c(1:22, "Un", "X", "Y")
row_ha <- rowAnnotation(chromosome = chr[keep], col = chrcols, na_col = "white")
column_ha <- HeatmapAnnotation(
  ethnicity = sampleInfo$ethnicity, group = factor(sampleInfo$group),
  col = list(
    ethnicity = c("ASN" = "lightgrey", "CEU" = "darkgrey"),
    group = c("0" = "brown", "1" = "blue")
  )
)
Heatmap(ge, use_raster = FALSE, top_annotation = column_ha, right_annotation = row_ha)
```

## Heatmaps: Summary

- Clustering becomes slow for thousands of rows but is great for visualizing co-expressed genes and sample groups.

## Colors

- Palettes: **sequential** (gradient), **diverging** (two directions from a center), **qualitative** (categorical).
- Keep color blindness in mind (10% of men). `RColorBrewer` has colorblind-friendly options.

## Plots To Avoid

> "Pie charts are a very bad way of displaying information." - R Help

- **Avoid pie charts and doughnut charts.** Humans are much better at judging length and position than angles and areas. A simple bar chart is almost always a better, clearer alternative.
- **Avoid pseudo 3D plots.** They distort the data and make comparisons difficult.
- Use color judiciously to highlight, not to decorate.

# Batch effects

## Pervasiveness of batch Effects

- Pervasive in genomics and have caused high-profile retractions.
- You can't get rid of them, but you can design your experiment to manage them.
- **The Golden Rule:** Make sure batch is not confounded with your variable of interest.
- Consider this **nightmare scenario**:

> * **Batch 1:** All your "Control" samples, processed on Monday.
> * **Batch 2:** All your "Treatment" samples, processed on Tuesday.
>
> If you find a difference, is it due to the treatment or the processing day? It's **impossible to know**.

Prevent such "confounding" through **blocking** and **randomization** during experimental design.

## Blocking and Randomization

* _Randomization_ is the process of randomly assigning participants or experimental units to different treatments or batches.
   - Randomization is the only way to guarantee there can be no systematic relationship between treatment and batch, or between study subject and treatment
* _Blocking_ is the process of grouping similar experimental units together to control for known sources of variability (e.g., time, technician, reagent lot).
   - For example, if you have multiple technicians, you can block by technician to ensure each treatment is applied by each technician.
   - This helps reduce variability and improve the accuracy of your results.
   - Only works for known sources of variability

## The batch effects impact clustering

```{r clust1}
# Data from a real study where date of processing was confounded with ethnicity
library(Biobase)
library(genefilter)
library(GSE5859)
data(GSE5859)
geneExpression <- exprs(e)
sampleInfo <- pData(e)
year <- as.integer(format(sampleInfo$date, "%y")) - min(as.integer(format(sampleInfo$date, "%y")))
hcclass <- cutree(hclust(as.dist(1 - cor(geneExpression))), k = 5)
table(hcclass, year) # Clustering is driven by year of processing, not biology
```

## Approaches to correcting for batch effects

Methods can be categorized by their approach and data type:

* **Simple Rescaling (e.g., `batchelor::rescaleBatches()`):**
    * Rescales batches to have the same mean/variance.
    * Good for single-cell data because it maintains sparsity (zeros stay zeros).

* **For Known Batches (Bulk RNA-seq):**
    * **`limma::removeBatchEffect()`** or **`sva::ComBat()`**: Use a linear model to regress out the effect of known batch variables (e.g., processing date, sequencing machine). Assumes cell type composition is similar across batches.

* **For Unknown Batches (Bulk RNA-seq):**
    * **`sva::sva()`**: Identifies and creates "surrogate variables" that capture hidden sources of variation. Excellent when you don't know the exact source of the batch effect but the p-value histogram looks problematic.

* **For Single-Cell Data Integration:**
    * **`batchelor::fastMNN()`**: Finds Mutual Nearest Neighbors (MNNs) between batches to align them. Powerful for single-cell RNA-seq because it does **not** assume the same composition of cells across batches.

## Batch Effects - summary

- Batch effects can ONLY be corrected if they are not confounded with your variable of interest.
- **Randomization is your best defense.**
- Always record info that might become a batch effect (date, technician, reagent lot, etc.).
- Include control samples in each batch.

## Links

-   A built [html](https://waldronlab.io/AppStatBio/articles/day4_batcheffects-vis.html)
    version of this lecture is available.
-   The [source](https://github.com/waldronlab/AppStatBio/blob/main/vignettes/day4_batcheffects-vis.Rmd) R Markdown is
     available from Github.

