---
title: "Lab 4: DAGs in Data Analysis"
author: "Levi Waldron"
output: html_document
---

## Introduction

In this lab, we will use simulated data to demonstrate the importance of Directed Acyclic Graphs (DAGs) in guiding statistical analysis. We will see how failing to control for a confounder, or mistakenly controlling for a collider, can both lead to biased estimates of a causal effect.

We will simulate a scenario where the true causal effect of an Exposure (E) on a Disease (D) is exactly zero. However, there are two other variables present:
- **C (Confounder):** Causes both E and D.
- **S (Collider):** Caused by both E and D.

Let's draw the DAG for this scenario:

```{r dag, message=FALSE, warning=FALSE}
library(ggdag)
library(ggplot2)

dag <- dagify(
  D ~ C,
  E ~ C,
  S ~ E + D,
  labels = c("D" = "Disease", "E" = "Exposure", "C" = "Confounder", "S" = "Collider"),
  exposure = "E",
  outcome = "D",
  coords = list(
    x = c(C = 0, E = -1, D = 1, S = 0),
    y = c(C = 1, E = 0, D = 0, S = -1)
  )
)

ggdag(dag, text = FALSE, use_labels = "label") + theme_dag()
```

## Data Simulation

We simulate 1000 observations based on the relationships defined in our DAG:

```{r simulate}
set.seed(123)
N <- 1000

# 1. C is a confounder (causes E and D)
C <- rnorm(N, mean = 0, sd = 1)

# 2. E is caused by C, with some random noise
E <- 0.5 * C + rnorm(N, mean = 0, sd = 1)

# 3. D is caused by C, with some random noise.
# Crucially, the true effect of E on D is 0.
D <- 0.5 * C + 0 * E + rnorm(N, mean = 0, sd = 1)

# 4. S is a collider (caused by both E and D)
S <- 0.5 * E + 0.5 * D + rnorm(N, mean = 0, sd = 1)

# Combine into a dataframe
data <- data.frame(E, D, C, S)
```

## Data Analysis

We want to estimate the causal effect of E on D. We know the true effect is 0. Let's see what happens under different modeling choices.

### 1. A Biased Model: Failing to Control for the Confounder

If we simply regress D on E without controlling for the confounder C, the path `E <- C -> D` remains open. This "backdoor path" creates a spurious association between E and D.

```{r model_biased_confounder}
summary(lm(D ~ E, data = data))
```

*Notice that the coefficient for E is statistically significant and positive, even though the true effect is 0!*

*Now the coefficient for E is very close to 0 and not statistically significant. We successfully recovered the true effect.*

### 2. Another Biased Model: Improperly Controlling for the Collider

What if we decide to throw all available variables into the model, controlling for both C and S? 

By conditioning on a collider (S), we open up a new non-causal path between E and D. This introduces "collider bias" (also known as Berkson's paradox or selection bias).

```{r model_biased_collider}
summary(lm(D ~ E + C + S, data = data))
```

*By controlling for the collider S, the coefficient for E becomes statistically significant and negative. We have introduced bias by over-adjusting!*

### 2. The Unbiased Model: Properly Controlling for the Confounder

To get an unbiased estimate of the effect of E on D, we must block the backdoor path by controlling for the confounder C. We must also be careful **not** to control for the collider S.

Try this yourself, show your code and output, and let me know what you find. What happens when you simply "control for everything"?

