---
title: "Day 2 mini-exercises"
author: "Levi Waldron"
date: "1/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load data

```{r, echo=FALSE}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
library(downloader)
if (!file.exists(filename))
    download(url, filename)
spider <- read.csv(filename, skip=1)
```

Using the `spider` object:

# Question 1. 

Create a new variable `leg2` where `L1` / `L2` are merged, and `L3` / `L4` are merged

```{r}
spider$leg2 <- spider$leg
levels(spider$leg)
levels(spider$leg2) <- c("L12", "L12", "L34", "L34")
with(spider, table(leg, leg2))
```

# Question 2. 

Create a new variable `leg3` that is an "ordered" factor (see `?factor`)

```{r}
spider$leg3 <- factor(spider$leg, ordered=TRUE)
head(spider$leg)
head(spider$leg3)
```

# Question 3. 

Make "push" the reference level for `spider$type` (see `?relevel`)

```{r}
spider$type <- relevel(spider$type, ref="push")
head(spider$type)
```

# Question 4

Perform regression on the spider dataset to model friction as a function of type, leg, *and* the interaction of type and leg. Interpret the output of this regression.

```{r}
fit <- lm(formula = friction ~ type * leg, data = spider)
summary(fit)
```

