---
title: "Population"
author: "Calogero Zarbo"
date: "February 23, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r preparation}
library(downloader) 
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- basename(url)
download(url, destfile=filename)
dat <- read.csv(filename) 
dat <- na.omit( dat )
```



1. Use dplyr to create a vector x with the body weight of all males on the control (chow) diet. What is this population’s average?

```{r ex1}
library(dplyr)
x <- filter(dat, Diet=='chow' & Sex=='M')$Bodyweight
x_mean <- mean(x)
print (x_mean)
```

2. Now use the rafalib package and use the popsd function to compute the population standard deviation.

```{r ex2}
library(rafalib)
x_sd <- popsd(x)
print (x_sd)
```

3. Set the seed at 1. Take a random sample X of size 25 from x. What is the sample average?

```{r ex3}
set.seed(1)
X <- sample(x,25)
X_mean <- mean(X)
print (X_mean)
```

4. Use dplyr to create a vector y with the body weight of all males on the high fat (hf) diet. What is this population’s average?

```{r ex4}
y <- filter(dat, Diet=='hf' & Sex=='M')$Bodyweight
y_mean <- mean(y)
print (y_mean)
```

5. Now use the rafalib package and use the popsd function to compute the population standard deviation.

```{r ex5}
y_sd <- popsd(y)
print (y_sd)
```

6. Set the seed at 1. Take a random sample Y of size 25 from y. What is the sample average?

```{r ex6}
set.seed(1)
Y <- sample(y,25)
Y_mean <- mean(Y)
print (Y_mean)
```

7. What is the difference in absolute value between $\bar{y}−\bar{x}$ and $\bar{X}-\bar{Y}$?

```{r ex7}
abs_diff <- (y_mean - x_mean) - (X_mean - Y_mean)
print (abs_diff)
```

8. Repeat the above for females. Make sure to set the seed to 1 before each sample call. What is the difference in absolute value between $\bar{y}−\bar{x}$ and $\bar{X}-\bar{Y}$?

```{r ex8}
x <- filter(dat, Diet=='chow' & Sex=='F')$Bodyweight
x_mean <- mean(x)
print (x_mean)
x_sd <- popsd(x)
print (x_sd)
set.seed(1)
X <- sample(x,25)
X_mean <- mean(X)
print (X_mean)
y <- filter(dat, Diet=='hf' & Sex=='F')$Bodyweight
y_mean <- mean(y)
print (y_mean)
y_sd <- popsd(y)
print (y_sd)
set.seed(1)
Y <- sample(y,25)
Y_mean <- mean(Y)
print (Y_mean)
abs_diff <- (y_mean - x_mean) - (X_mean - Y_mean)
print (abs_diff)
```

9. For the females, our sample estimates were closer to the population difference than with males. What is a possible explanation for this?

    A) The population variance of the females is smaller than that of the males; thus, the sample variable has less variability.
    B) Statistical estimates are more precise for females.
    C) The sample size was larger for females.
    D) The sample size was smaller for females.


```{r ex9}
print ("A) The population variance of the females is smaller than that of the males; thus, the sample variable has less variability.")

```

