---
title: "Power Calculations Exercises"
author: "Francesco Asnicar"
date: "24 February 2016"
output: html_document
---

For these exercises we will load the babies dataset from babies.txt. We will use this data to review the concepts behind the p-values and then test confidence interval concepts.

```{r}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/babies.txt"
filename <- basename(url)
if (!file.exists(filename)) {
  download(url, destfile=filename)
}
babies <- read.table("babies.txt", header=TRUE)
```

This is a large dataset (1,236 cases), and we will pretend that it contains the entire population in which we are interested. We will study the differences in birth weight between babies born to smoking and non-smoking mothers.

First, let’s split this into two birth weight datasets: one of birth weights to non-smoking mothers and the other of birth weights to smoking mothers.

```{r}
library(dplyr)
bwt.nonsmoke <- filter(babies, smoke==0) %>% select(bwt) %>% unlist 
bwt.smoke <- filter(babies, smoke==1) %>% select(bwt) %>% unlist
```

Now, we can look for the true population difference in means between smoking and non-smoking birth weights.

```{r}
library(rafalib)
mean(bwt.nonsmoke)-mean(bwt.smoke)
popsd(bwt.nonsmoke)
popsd(bwt.smoke)
```

The population difference of mean birth weights is about 8.9 ounces. The standard deviations of the nonsmoking and smoking groups are about 17.4 and 18.1 ounces, respectively.

As we did with the mouse weight data, this assessment interactively reviews inference concepts using simulations in R. We will treat the babies dataset as the full population and draw samples from it to simulate individual experiments. We will then ask whether somebody who only received the random samples would be able to draw correct conclusions about the population.

We are interested in testing whether the birth weights of babies born to non-smoking mothers are significantly different from the birth weights of babies born to smoking mothers.

#1
Set the seed at `1` and obtain two samples, each of size `N=25`, from non-smoking mothers (`dat.ns`) and smoking mothers (`dat.s`). Compute the t-statistic (call it `tval`).

```{r}
set.seed(1)
dat.ns <- sample(bwt.nonsmoke, size=25)
set.seed(1)
dat.s <- sample(bwt.smoke, size=25)
tval <- t.test(dat.s, dat.ns)$statistic
```

#2
Recall that we summarize our data using a t-statistics because we know that in situations where the null hypothesis is true (what we mean when we say “under the null”) and the sample size is relatively large, this t-value will have an approximate standard normal distribution. Because we know the distribution of the t-value under the null, we can quantitatively determine how unusual the observed t-value would be if the null hypothesis were true.

The standard procedure is to examine the probability a t-statistic that actually does follow the null hypothesis would have larger absolute value than the absolute value of the t-value we just observed – this is called a two-sided test.

We have computed these by taking one minus the area under the standard normal curve between `-abs(tval)` and `abs(tval)`. In R, we can do this by using the `pnorm` function, which computes the area under a normal curve from negative infinity up to the value given as its first argument:

```{r}
-abs(tval)
abs(tval)
1 - pnorm(abs(tval)) - pnorm(-abs(tval))
```

#3
Because of the symmetry of the standard normal distribution, there is a simpler way to calculate the probability that a t-value under the null could have a larger absolute value than tval. Choose the simplified calculation from the following:

* A) `1-2*pnorm(abs(tval))`
* B) `1-2*pnorm(-abs(tval))`
* C) `1-pnorm(-abs(tval))`
* D) `2*pnorm(-abs(tval))`

ANSWER: D

```{r}
2*pnorm(-abs(tval))
```

#4
By reporting only p-values, many scientific publications provide an incomplete story of their findings. As we have mentioned, with very large sample sizes, scientifically insignificant differences between two groups can lead to small p-values. Confidence intervals are more informative as they include the estimate itself. Our estimate of the difference between babies of smoker and non-smokers: `mean(dat.s) - mean( dat.ns)`. If we use the CLT, what quantity would we add and subtract to this estimate to obtain a 99% confidence interval?

```{r}
diff.means <- mean(dat.s) - mean(dat.ns)
se <- sd(c(dat.s, dat.ns))/sqrt(length(dat.s)+length(dat.ns))
c(diff.means-2.58*se, diff.means+2.58*se)
```

#5
#6
#7
#8
#9
#10
#11
#12