---
title: "Population, Samples, and Estimates Exercises"
author: "Serena"
date: "22 febbraio 2016"
output: html_document
---


```{r}
library(downloader)
library(dplyr)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- basename(url)
download(url, destfile=filename)
dat <- read.csv(filename) 

dat <- na.omit( dat )
```
#1. Male control population average
```{r}
x <- dplyr::filter(dat, dat$Sex == 'M' & dat$Diet == "chow")
meanx <- mean(x$Bodyweight)
meanx
```
#2. Male control population standard deviation
```{r}
rafalib::popsd(x$Bodyweight)
```
#3. Sample average from x with seed=1 and size=25
```{r}
set.seed(1)
smeanx <- mean(sample(x$Bodyweight, size=25))
smeanx
```
#4. Male high-fat population average
```{r}
y <- dplyr::filter(dat, dat$Sex =='M' & dat$Diet=="hf")
meany <- mean(y$Bodyweight)
meany
```
#5. Male high-fat population standard deviation
```{r}
rafalib::popsd(y$Bodyweight)
```
#6. Sample average from y with seed=1 and size=25
```{r}
set.seed(1)
smeany <- mean(sample(y$Bodyweight, size=25))
```
#7. What is the difference in absolute value?
### Population means
```{r}
abs(meany-meanx)
```
### Sample means
```{r}
abs(smeany-smeanx)
```
#8. Females
### Female control population average and standard deviation
```{r}
xf <- dplyr::filter(dat, dat$Sex == 'F' & dat$Diet == "chow")
meanxf <- mean(xf$Bodyweight)
meanxf

rafalib::popsd(xf$Bodyweight)
```
### Sample average from x with seed=1 and size=25
```{r}
set.seed(1)
smeanxf <- mean(sample(xf$Bodyweight, size=25))
smeanxf
```
### Female high-fat population average and standard deviation
```{r}
yf <- dplyr::filter(dat, dat$Sex =='F' & dat$Diet=="hf")
meanyf <- mean(yf$Bodyweight)
meanyf

rafalib::popsd(yf$Bodyweight)
```
### Sample average from y with seed=1 and size=25
```{r}
set.seed(1)
smeanyf <- mean(sample(yf$Bodyweight, size=25))
```
### Population means
```{r}
abs(meanyf-meanxf)
```
### Sample means
```{r}
abs(smeanyf-smeanxf)
```
#9. Females sample estimates
### For the females, our sample estimates were closer to the population difference than with males
###A) The population variance of the females is smaller than that of the males; thus, the sample variable has less variability.

