---
title: "Distance"
author: "Francesco Asnicar"
date: "10 March 2016"
output: 
  html_document: 
    number_sections: yes
    toc: yes
---

If you have not done so already, install the data package tissueGeneExpression:

```{r}
library(devtools)
## install_github("genomicsclass/tissuesGeneExpression")
# install_github("genomicsclass/tissuesGeneExpression", force=T)
```

The data represents RNA expression levels for eight tissues, each with several *biological replictes*. We call samples that we consider to be from the same population, such as liver tissue from different individuals, *biological replicates*:

```{r}
library(tissuesGeneExpression)
data(tissuesGeneExpression)
head(e)
head(tissue)
```

#1
How many biological replicates for hippocampus?

```{r}
length(which(tissue == "hippocampus"))
```

#2
What is the distance between samples 3 and 45?

```{r}
dist( t( e[, c(3,45)] ) )
```

#3
What is the distance between gene `210486_at` and `200805_at`

```{r}
dist( e[c("210486_at", "200805_at"),] )
```

#4
If I run the command (don’t run it!):

```{r}
# d = as.matrix( dist(e) )
```

how many cells (number of rows times number of columns) will this matrix have?

ANSWER: `nrow(e)*nrow(e)` = `r nrow(e)*nrow(e)`

#5
Compute the distance between all pair of samples:

```{r}
d = dist( t(e) )
```

Read the help file for `dist`. How many distances are stored in `d`? Hint: What is the length of `d`?

ANSWER: `length(d)` = `r length(d)`

#6
Why is the answer to exercise 5 not `ncol(e)^2`?

* A) R made a mistake there.
* B) Distances of 0 are left out.
* C) Because we take advantage of symmetry: only lower triangular matrix is stored thus only `ncol(e)*(ncol(e)-1)/2` values.
* D) Because it is equal `nrow(e)^2`

ANSWER: C
