---
title: "MDS"
author: "Calogero Zarbo"
date: "April 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using the z we computed in exercise 4 of the previous exercises.
Build mds plot
Use `cmdscale` on the original data
What is the absolute value of the correlation between the first dimension of z and the first dimension in mds?



```{r}
library(tissuesGeneExpression)
data(tissuesGeneExpression)
y = e - rowMeans(e)
s = svd(y)
z = s$d * t(s$v)

library(rafalib)
ftissue = factor(tissue)
mypar(1, 1)
plot(z[1,], z[2,], col=as.numeric(ftissue))
legend("topleft", levels(ftissue), col=seq_along(ftissue), pch=1)

d = dist(t(e))
mds = cmdscale(d)

abs(cor(z[1,], mds[,1]))
```

What is the absolute value of the correlation between the second dimension of z and the second dimension in mds?

```{r}
abs(cor(z[2,], mds[,2]))
```


Load the following dataset:

```{r}
library(devtools)
install_github("genomicsclass/GSE5859Subset")
library(GSE5859Subset)
data(GSE5859Subset)
```

Compute the svd and compute z.
Which dimension of `z` most correlates with the outcome `sampleInfo$group`?

```{r}
s = svd(geneExpression-rowMeans(geneExpression))
z = s$d * t(s$v)

which.max(cor(t(z), sampleInfo$group))
```


What is this max correlation?

```{r}
max(cor(t(z), sampleInfo$group))
```


Which dimension of `z` has the second highest correlation with the outcome `sampleInfo$group`?

```{r}
max.second <- sort(cor(t(z), sampleInfo$group), decreasing=T)[2]
which(cor(t(z), sampleInfo$group) == max.second)
```


Note these measurements were made during two months:

```{r}
sampleInfo$date
```

We can extract the month this way:

```{r}
month = format(sampleInfo$date, "%m")
month = factor(month)
```

Which dimension of `z` has the second highest correlation with the outcome `month`?

```{r}
second.month <- which(cor(t(z), as.numeric(month)) == sort(cor(t(z), as.numeric(month)), decreasing=T)[2])
second.month
```


What is this correlation?

```{r}
cor(t(z), as.numeric(month))[second.month]
```
